Synthesising Recursive Functions for First-Order Model Counting:\\ Challenges, Progress, and Conjectures

### 2023-03-07 doc/paper/paper.tex
Clean up supplementary material

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize]{cleveref}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{complexity}
\usepackage{siunitx}
\usepackage{forest}
\usepackage[inline]{enumitem}
\usepackage{subcaption}

@@

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{conjecture}{Conjecture}
\newtheorem{claim}{Claim}[theorem]
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\setlength{\multlinegap}{0pt}

@@

% TODO (at the very end)
% grammarly the whole thing
% read the formatting instructions and their LaTeX source (email font?)
% make sure there are no overflow errors
% make sure the result table (as well as other floating objects) are on the right pages

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand{\Done}{\mathsf{D}_{1}}
\newcommand{\Dtwo}{\mathsf{D}_{2}}
\newcommand{\Dthree}{\mathsf{D}_{3}}

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\GDR$ nodes.}\label{alg:domainrecursion}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{domain $\Omega \in \mathcal{D}$ s.t.\ there is $c \in \phi$ and $X \in \Vars(L_c)$ s.t. $\delta_c(X) = \Omega$\label{line:condition}}{
    $\phi' \gets \emptyset$\;
    $x \gets \text{a new constant in domain } \Omega$\;\label{line:constant}
    \ForEach{clause $c = (L, E, \delta) \in \phi$\label{line:forclause}}{
      $V \gets \{\, X \in \Vars(L) \mid \delta(X) = \Omega \,\}$\;\label{line:V}
      \ForEach{$W \subseteq V$ s.t. $W^2 \cap E = \emptyset$ {\bf and} $W \cap \{\, X \in \Vars(E) \mid (X, y) \in E \text{ for some constant } y \,\} = \emptyset$\label{line:conditions}}{
        \tcc{$\delta'$ restricts $\delta$ to the new set of variables}
        $\phi' \gets \phi' \cup \{\, (L[x/W], E[x/W] \cup \{\, (X, x) \mid (X \in V \setminus W) \,\}, \delta') \,\}$\;\label{line:generation}
      }
    }
    $S \gets S \cup \{\, (\GDR, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\CR$ nodes.}\label{alg:constraintremoval}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{replaceable pair $(\Omega \in \mathcal{D}, x \in \Omega)$\label{line:crconditions}}{
    add a new domain $\Omega'$ to $\mathcal{D}$\;\label{line:newdomain}
    $\phi' \gets \emptyset$\;
    \ForEach{clause $(L, E, \delta) \in \phi$}{
      $E' \gets \{\, (a, b) \in E \mid b \ne x \,\}$\;\label{line:constraintremoval}
      \nosemic$\delta' \gets X \mapsto
      \begin{cases}
        \Omega' & \text{if } \delta(X) = \Omega\\
        \delta(X) & \text{otherwise;}
      \end{cases}$\;\label{line:newdelta}
      $\phi' \gets \phi' \cup \{\, (L, E', \delta') \,\}$\;
    }
    $S \gets S \cup \{\, (\CR_{\Omega \mapsto \Omega'}, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

\paragraph{Hashing.}
We use (integer-valued) hash functions to discard pairs of formulas that are too
different for recursion to work. The hash code of a clause
$c = (L, E, \delta_{c})$ (denoted by $\# c$) combines the hash codes of the sets
of constants and predicates in $c$, the numbers of positive and negative
literals, the number of inequality constraints $|E|$, and the number of
variables $|\Vars(c)|$. The hash code of a formula $\phi$ combines the hash
codes of all its clauses and is denoted by $\#\phi$.

\paragraph{Caching.}
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} uses a cache to check if
a formula is identical to one of the formulas that have already been fully
compiled. To facilitate recursion, we extend the caching scheme to include
formulas that have been encountered but not fully compiled yet. Formally, we
define a \emph{cache} to be a map from integers (e.g., hash codes) to sets of
pairs of the form $(\phi, v)$, where $\phi$ is a formula, and $v$ is an FCG
node.

\begin{algorithm}[t]
  \caption{The compilation rule for $\Reff$ nodes.}\label{alg:trycache}
  \KwIn{formula $\phi$, cache $C$}
  \KwOut{a set of chips}

@@

\paragraph{Miscellaneous other instances.}
We also note that standard (W)FOMC benchmarks such as \cref{example:smokers}
that are already supported by \textsc{ForcLift} remain feasible for
\textsc{Crane} as well. Moreover, our algorithm can also handle the linear order
axiom, for which a bespoke algorithm was recently proposed by T{\'{o}}th and
Ku\v{z}elka~\shortcite{DBLP:journals/corr/abs-2211-01164}. To demonstrate that
\textsc{Crane} is capable of finding efficient solutions to problems beyond
known domain-liftable fragments such as \Ctwo{}, we run \textsc{Crane} in greedy
search mode on the following formula:
\begin{equation}\label[formula]{eq:ternary}
  \begin{split}
    &\begin{multlined}[t]
      (\forall X,W \in \Gamma\text{. }\forall Y \in \Delta\text{. }\forall Z \in \Lambda\text{. }\\
      \texttt{p}(X, Y, Z) \land \texttt{p}(W, Y, Z) \Rightarrow X = W) \land{}
    \end{multlined}\\
    &\begin{multlined}[t]
      (\forall X \in \Gamma\text{. }\forall Y,W \in \Delta\text{. }\forall Z \in \Lambda\text{. }\\
      \texttt{p}(X, Y, Z) \land \texttt{p}(X, W, Z) \Rightarrow Y = W).
    \end{multlined}
  \end{split}
\end{equation}
For every element $z$ of $\Lambda$, \cref{eq:ternary} restricts
$\texttt{p}(X, Y, z)$ to be a $\Gamma \to \Delta$ partial injection, independent
of $\texttt{p}(X, Y, z')$ for any $z' \ne z$. \textsc{Crane} instantly returns a
solution of complexity $\Theta(l+mn)$, where $l = |\Lambda|$.

\section{Conclusion and Future Work}\label{sec:conclusion}

@@

\bibliographystyle{kr}
\bibliography{paper}

>>>>

\usepackage{mathtools}
\usepackage{amssymb}
\usepackage{pifont}
\usepackage[linesnumbered,ruled,vlined]{algorithm2e}
\usepackage{tikz}
\usepackage[capitalize,noabbrev]{cleveref}
\usepackage{multirow}
\usepackage{colortbl}
\usepackage{complexity}
\usepackage{siunitx}
\usepackage{forest}
\usepackage[inline]{enumitem}
\usepackage{subcaption}

@@

% See https://www.overleaf.com/learn/latex/theorems_and_proofs
% for a nice explanation of how to define new theorems, but keep
% in mind that the amsthm package is already included in this
% template and that you must *not* alter the styling.
% \newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{fact}{Fact}
\newtheorem{proposition}{Proposition}
\newtheorem{conjecture}{Conjecture}
\newtheorem{claim}{Claim}[theorem]
\newtheorem{example}{Example}
\newtheorem{definition}{Definition}

@@

% TODO (at the very end)
% grammarly the whole thing (supplementary material as well)
% make sure the result table (as well as other floating objects) are on the right pages
% make sure the abstract, introduction, and conclusion are all in line with each other
% maybe improve the indentation of equations

@@

\newcommand{\crefrangeconjunction}{--}
\newcommand\pfun{\mathrel{\ooalign{\hfil$\mapstochar\mkern5mu$\hfil\cr$\to$\cr}}}
\newcommand{\cmark}{\ding{51}}%
\newcommand{\xmark}{\ding{55}}%
\newcommand{\FOtwo}{$\mathsf{FO}^{2}$}
\newcommand{\FOthree}{$\mathsf{FO}^{3}$}
\newcommand{\SFO}{$\mathsf{S}^{2}\mathsf{FO}^{2}$}
\newcommand{\SRU}{$\mathsf{S}^{2}\mathsf{RU}$}
\newcommand{\Uone}{$\mathsf{U}_{1}$}
\newcommand{\Ctwo}{$\mathsf{C}^{2}$}
\newcommand{\IFO}{$\mathsf{I}\mathsf{FO}^{2}$}
\newcommand\mdoubleplus{\mathbin{+\mkern-10mu+}}
\newcommand{\Done}{\mathsf{D}_{1}}
\newcommand{\Dtwo}{\mathsf{D}_{2}}
\newcommand{\Dthree}{\mathsf{D}_{3}}
\newcommand{\predicate}{\texttt{\textup{p}}}

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\GDR$ nodes}\label{alg:domainrecursion}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{domain $\Omega \in \mathcal{D}$ s.t.\ there is $c \in \phi$ and $X \in \Vars(L_c)$ s.t. $\delta_c(X) = \Omega$\label{line:condition}}{
    $\phi' \gets \emptyset$\;
    $x \gets \text{a new constant in domain } \Omega$\;\label{line:constant}
    \ForEach{clause $c = (L, E, \delta) \in \phi$\label{line:forclause}}{
      $V \gets \{\, X \in \Vars(L) \mid \delta(X) = \Omega \,\}$\;\label{line:V}
      \ForEach{$W \subseteq V$ s.t. $W^2 \cap E = \emptyset$ {\bf and} $W \cap \{\, X \in \Vars(E) \mid (X, y) \in E \text{ for some constant } y \,\} = \emptyset$\label{line:conditions}}{
        \tcc{$\delta'$ restricts $\delta$ to the new set of variables}
        $\phi' \gets \phi' \cup \{\, (L[x/W], E[x/W] \cup \{\, (X, x) \mid (X \in V \setminus W) \,\}, \delta') \,\}$\;\label{line:generation}
      }
    }
    $S \gets S \cup \{\, (\GDR, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\CR$ nodes}\label{alg:constraintremoval}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{replaceable pair $(\Omega \in \mathcal{D}, x \in \Omega)$\label{line:crconditions}}{
    add a new domain $\Omega'$ to $\mathcal{D}$\;\label{line:newdomain}
    $\phi' \gets \emptyset$\;
    \ForEach{clause $(L, E, \delta) \in \phi$}{
      $E' \gets \{\, (a, b) \in E \mid b \ne x \,\}$\;\label{line:constraintremoval}
      \nosemic$\delta' \gets X \mapsto
      \begin{cases}
        \Omega' & \text{if } \delta(X) = \Omega\\
        \delta(X) & \text{otherwise;}
      \end{cases}$\;\label{line:newdelta}
      $\phi' \gets \phi' \cup \{\, (L, E', \delta') \,\}$\;
    }
    $S \gets S \cup \{\, (\CR_{\Omega \mapsto \Omega'}, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

\subsubsection{Hashing}
We use (integer-valued) hash functions to discard pairs of formulas that are too
different for recursion to work. The hash code of a clause
$c = (L, E, \delta_{c})$ (denoted by $\# c$) combines the hash codes of the sets
of constants and predicates in $c$, the numbers of positive and negative
literals, the number of inequality constraints $|E|$, and the number of
variables $|\Vars(c)|$. The hash code of a formula $\phi$ combines the hash
codes of all its clauses and is denoted by $\#\phi$.

\subsubsection{Caching}
\textsc{ForcLift}~\cite{DBLP:conf/ijcai/BroeckTMDR11} uses a cache to check if
a formula is identical to one of the formulas that have already been fully
compiled. To facilitate recursion, we extend the caching scheme to include
formulas that have been encountered but not fully compiled yet. Formally, we
define a \emph{cache} to be a map from integers (e.g., hash codes) to sets of
pairs of the form $(\phi, v)$, where $\phi$ is a formula, and $v$ is an FCG
node.

\begin{algorithm}[t]
  \caption{The compilation rule for $\Reff$ nodes}\label{alg:trycache}
  \KwIn{formula $\phi$, cache $C$}
  \KwOut{a set of chips}

@@

\subsubsection{Miscellaneous Other Instances}
We also note that standard (W)FOMC benchmarks such as \cref{example:smokers}
that are already supported by \textsc{ForcLift} remain feasible for
\textsc{Crane} as well. Moreover, our algorithm can also handle the linear order
axiom, for which a bespoke algorithm was recently proposed by T{\'{o}}th and
Ku\v{z}elka~\shortcite{DBLP:journals/corr/abs-2211-01164}. To demonstrate that
\textsc{Crane} is capable of finding efficient solutions to problems beyond
known domain-liftable fragments such as \Ctwo{}, we run \textsc{Crane} in greedy
search mode on the following formula:
\begin{equation}\label[formula]{eq:ternary}
  \begin{split}
    &\begin{multlined}[t]
      (\forall X,W \in \Gamma\text{. }\forall Y \in \Delta\text{. }\forall Z \in \Lambda\text{. }\\
      \texttt{p}(X, Y, Z) \land \texttt{p}(W, Y, Z) \Rightarrow X = W) \land{}
    \end{multlined}\\
    &\begin{multlined}[t]
      (\forall X \in \Gamma\text{. }\forall Y,W \in \Delta\text{. }\forall Z \in \Lambda\text{. }\\
      \texttt{p}(X, Y, Z) \land \texttt{p}(X, W, Z) \Rightarrow Y = W).
    \end{multlined}
  \end{split}
\end{equation}
For every element $z$ of $\Lambda$, \cref{eq:ternary} restricts
$\texttt{p}(X, Y, z)$ to be a $\Gamma \to \Delta$ partial injection, independent
of $\texttt{p}(X, Y, z')$ for any $z' \ne z$. \textsc{Crane} instantly returns a
solution of complexity $\Theta(l+mn)$, where $l = |\Lambda|$.

\section{Conclusion and Future Work}\label{sec:conclusion}
\setlength{\multlinegap}{0pt}

@@

% TODO: The reference list should not include unpublished work. (earnest + OEIS + 3 corr references)
% TODO: replace all instances of {p} with \predicate (by hand)
\bibliographystyle{kr}
\bibliography{paper}
